\sysfull

### 2017-07-28 paper/background.tex
Background revised almost

\sys addresses the problem of intermittently executing task-based computations
on energy-harvesting hardware, without being limited by inflexible, static task
decomposition and without being limited to fully non-volatile main memory.  In
this section, we provide background on energy harvesting systems and the
intermittent software execution model.   We then discuss the key limitations of
existing task-based programming and execution models for intermittent computing
that \sys addresses.

@@

Batteryless operation has a number of important advantages, making intermittent
computing an important domain.  Supplying power to the ``next
billion''~\cite{citationneeded} embedded computers using batteries is not
sustainable.  The European Commission estimates that more than 160 kilotons of
consumer batteries enter the European Union annually~\cite{eu_batteries_2016}.
Batteries constitute a potential environmental hazard because not all batteries
are recycled.  Batteries are fragile and limited in their number of
charge/discharge cycles, requiring costly maintenance that is possibly only
through physical device access, which may be difficult if devices are deployed
in inaccessible places (e.g., in space~\cite{kicksat}). By contrast,
super-capacitors are durable, promising millions of charge/discharge
cycles~\cite[Sec. I]{ongaro_pwre_2012} The main limitation in moving to
capacitor-based energy storage is that capacitor energy density --- and
consequently operating discharge time --- is orders of magnitude less than a
battery.

\todo{Compress text}{Brandon}
Given current technology development, battery-less systems are best suited for
very long-term sensing and monitoring where access to recharge is either
prohibitive or impossible. These include battery-less image capture and
processing~\cite{naderiparizi_rfid_2015}, animal monitoring~\cite{thomas_jbcs_2012} or implantable~\cite{rodriguez_tbcs_2015}
and digestible~\cite{nadeau_naturebio_2017} sensors.

@@

\begin{table}
	\centering
	\begin{tabular}{|c|c|}
	\hline
	Platform name & Storage capacitor size \\
	\hline \hline
	Moo~\cite{moo} & 0.1\,F \\
	WISPCam~\cite{naderiparizi_rfid_2015} & 6.08\,mF \\ %tested [11.24, 17.45, 21.98]\,mF
	NFC-WISP~\cite{zhao_rfid_2015} & 300\,$\mu$F \\
	NeuralWISP~\cite{holleman_biocas_2008} & 100\,$\mu$F \\
	WISP~\cite{wisp5} & 47\,$\mu$F \\
	{\em BioImpedance} sensor~\cite{rodriguez_tbcs_2015} & 20\,$\mu$F \\
	{\em Ingestible} sensor~\cite{nadeau_naturebio_2017} & 220\,$\mu$F\\
	\hline
	\end{tabular} 
\caption{Comparison of {\em default} energy storage sizes for various battery-less platforms. Observe a huge variation in storage capacity. We note that values for other representative platforms~\cite{medusa_farsens_2017,talla_imwut_2017,liu_sigcomm_2013,parks_sigcomm_2014} were not reported in their respective papers.}
\label{table:capacitor}
\end{table}

@@

Before proceeding further with the background we need to make certain assumptions about the underlying hardware. We assume that we operate on the microcontroller (more generally, computing hardware) equipped with a mix of volatile and non-volatile memory, such as FLASH, Ferroelectric RAM or Magnetoresistive RAM. \sys does not assume to rely on specific memory type (either volatile or non-volatile), in contrary to earlier works targeting non-volatile processors~\cite{su_date_2017,ratchet,quickrecall,nvp}. Actually, \sys will exploit the properties of both memory types to maximize performance metrics of the program being executed. We make \sys able to I/O operations, however we do not enable the scheduling of operations (which \emph{nota bene} forms a foundation for a simple runtime). All in all, our hardware model follows in assumptions the two most relevant intermittently-powered execution environments~\cite{alpaca,chain}.

@@

We can classify the intermittent execution models into two classes: \emph{sequential execution model} (which historically formed a baseline for this stream of research) and \emph{modular execution model} (a potent alternative, on which we base the design decisions of \sys).
)
\subsubsection{Baseline---Sequential Execution Model}

Under this model a program for intermittently-powered device is seen as a one idempotent operational region with one common context. Generally, the progress of the program is saved and updated by means of checkpointing, refer again to Section~\ref{sec:background_consistency}. Normally, the sequential model relies on a hardware assistant to measure the voltage level in the energy reservoir to place a checkpoint~\cite{mementos,mottola2017harvos,hibernus}. The benefit of this model is that it does not require code modification by a programmer. However it suffers from significant overhead (refer to experiment result in~\cite[Fig. 3]{chain}, where it is concluded that each checkpoint causes overhead).

@@

At the heart of this model is the concept of an \emph{idempotent task}. The idempotent task is a C language function that does not have arguments and does not return a value. This task uses a well defined interface to interact with the non-volatile memory. Therefore, it tolerates arbitrary number of power interrupts. This model, generally, produces less overhead~\cite{chain} and allows multiple applications to run on an intermittently-powered device by interleaving their tasks~\cite{}. However, it obviously requires code modification. For example, if an algorithm is written according to the continuous execution model it has to be split by a programmer, into small tasks to run under the modular intermittent execution model.

Despite the benefit of the modular model certain limitations are still present that we must overcome and which inspired us to design \sys. Let us look at these problems in detail.

\begin{table}
	\centering
	\begin{tabular}{|c|c|c|c|}
		\hline
		{~} & CP & SZ & NV \\
		\hline\hline
		\!\!Mementos~\cite{mementos}\!\! & \!\!$N_2\leq N_1$\!\! & \!\!R + S\!\! & \!\!R + S\!\! \\
		\!\!DINO~\cite{dino}\!\! & $N_2$\!\! & \!\!R + W\!\! & \!\!R + W\!\! \\
		\!\!Chain~\cite{chain}, Alpaca~\cite{alpaca}\!\! & \!\!$N_2$\!\! & P\!\! & \!\!2W + P\!\!\\
		\!\!Ratchet~\cite{ratchet}, Clank~\cite{hicks_isca_2017}\!\! & $N_3\geq N_2\geq N_1$\!\! & \!\!R\!\! & R\!\!\\
		\hline 
	\end{tabular}
	\caption{Cost of various checkpointing methods; Symbols---R: register, S: stack, W: WAR-dependent variables, P: program counter, $N_1, N_2, N_3$: relative number of checkpoints for each method, CP: no. checkpoints, SZ: checkpoint size, NV: copy to non-volatile memory size.}
	\label{table:chechpoint_comparison}
\end{table}

@@

We have executed an experiment to measure exactly cost of read/write operations to volatile and non-volatile memory. The result is given in Figure~\ref{fig:framEnergy}. Four applications were developed to measure the energy consumption of \emph{accessing} the FRAM and SRAM of the MSP430FR5969 microcontroller~\cite{msp430datasheet}. The interference free debugger (EDB)~\cite{edb} is used as the energy measurement tool. The EDB probes the energy buffer before and after accessing the FRAM/SRAM 1600 times. This large number of FRAM access is used to increase the reliability of the results generated by the used measurement tool (e.g. reducing the effect of the quantization error). The energy buffer is charged, using the EDB, to $\approx$ 2.45\,V \textbf{only} at the beginning of the execution and of the programs. The write operations were performed by writing literal values, random numbers, to the memory. However, a read operation must be followed by a write operation. Therefore, we chose to write a read value to SRAM. Based on our measurements we conclude that a FRAM access equals $\frac{3}{2}$ SRAM access~\footnote{We would like to mention that TI has already mentioned in this document~\cite{} that FRAM write access is more energy expensive than SRAM write access. However, TI does not quantify the energy overhead.}. Therefore, reducing FRAM access is desirable when developing energy limited software solution---\emph{which is in sharp contrast to the main goal of all checkpoining solutions}.

The above result reassures the intuitive observation that a decrease of CP (Table~\ref{table:chechpoint_comparison}) decreases the energy cost associated with each state preservation method and reduces operation cost which in most simplistic terms is defined as $\mathcal{O}(\text{CP})+\mathcal{O}(\text{PL})$, where PL is the cost of program execution without checkpointing. On the other hand, this increases task/checkpoint re-execution time. The ideal solution is a runtime that \emph{adaptively} changes task size during runtime, which brings us to a second problem.

@@

A modular intermittent execution model completely depends on a programmer's estimation which mostly results in a sub-optimal code division. Moreover, this model can only consider a single hardware configuration and it does not take environment changing into considerations. Each battery-less platform has different storage sizes, refer to Table~\ref{table:capacitor}, which causes different charge/discharge periods. This result in different execution times and in the worst case the size of the capacitor might be too small to execute one task written by the programmer. This calls for the ability of the modular execution runtimes to \textbf{merge} multiple small tasks into a larger one or \textbf{split} a large task into a set of small ones on the fly (post-compilation time). Unfortunately the following features prevent us to perform this task using existing runtimes~\cite{chain,alpaca}.

@@

\begin{itemize}
	\item \textbf{Lack of multi-versioning system} All known modular execution models define its own input for each task that is not shared with other tasks. This design decision makes the benefit of merging tasks of less importance since each task is interacting with its own variables.

	\item \textbf{Irreversible tasks transitions} According to both Chain~\cite{chain} and Alpaca~\cite{alpaca} programming models, each task calls the next one. The transition is done through functions calls and manually clearing the stack, to prevent the stack overflow problem. This approach makes all the task transitions firm, preventing task merging.

	\item \textbf{`All channels' requirement} Fig.~\ref{fig:DynamicChainSeq} shows that if the task control flow is a circular with no branches then task merging is possible. Moreover, all the intra-merged-tasks channel can be committed to SRAM instead of FRAM to save energy. However, this execution path is only a special one. If the general case (see Fig.~\ref{fig:DynamicChainRan}) is considered then leaving out the intra-merged-task channels might result in data inconsistency. Than is, if an application tries to access a merged task from the middle after a power interrupt then the input for that task is not ready, which result in an incorrect execution.

	\item \textbf{Dependency on Privatization} Alpaca~\cite{alpaca} extended Chain~\cite{chain} with a concept of privatization, i.e. it protects task-shared variables from WAR dependency problem \emph{within} the scope of a task. Therefore, if tasks are on-demand merged the scope of these dependencies are changed and the static analysis to privatize these variables is not valid anymore and data consistency can not be preserved.
\end{itemize}

@@

\begin{table}
	\begin{tabular}{|c|c|c|c|}
		\hline
		{~} & Alpaca~\cite{alpaca} & Chain~\cite{chain} & This work \\
		\hline\hline
		Task connections & in-task & in-task & outside \\
		Task re-use & N & N & Y\\
		Memory model & globals & globals & globals\\
		Concurrency & N & N & N \\
		Task blocking & N & N & Y \\
		Energy-adaptive & N & N & Y \\
		\hline
	\end{tabular}
	\caption{Comparison of existing task-based execution models against \sys.}
	\label{table:feature_comparison}
\end{table}

@@

%%%Notes%%%

>>>>

\sys addresses the problem of intermittently executing task-based computations on energy-harvesting hardware, without being limited by inflexible, static task decomposition and without being limited to fully non-volatile main memory. In this section, we provide background on energy harvesting systems and the intermittent software execution model. We then discuss the key limitations of existing task-based execution models for intermittent computing that \sys addresses.

@@

Batteryless operation has a number of important advantages, making intermittent computing an important research domain. Supplying power to billions~\cite{gartner_iot} of embedded computers using batteries is not sustainable. The European Commission estimates that more than 160 kilotons of consumer batteries enter the European Union annually~\cite{eu_batteries_2016}. Batteries constitute a potential environmental hazard because not all batteries are recycled. Batteries are fragile and limited in their number of charge/discharge cycles, requiring costly maintenance that is possibly only through physical device access, which may be difficult if devices are deployed in inaccessible places, e.g. in space~\cite{kicksat}. By contrast, super-capacitors are durable, promising millions of charge/discharge cycles~\cite[Sec. I]{ongaro_pwre_2012}. The main limitation in moving to capacitor-based energy storage is that capacitor energy density---and consequently: operating discharge time---is orders of magnitude less than a battery. \todo{Compress text}{Brandon}

Given current technology development, battery-less systems are best suited for very long-term sensing and monitoring where access to recharge is either prohibitive or impossible. These include battery-less image capture and processing~\cite{naderiparizi_rfid_2015}, animal monitoring~\cite{thomas_jbcs_2012} or implantable~\cite{rodriguez_tbcs_2015} and digestible~\cite{nadeau_naturebio_2017} sensors.

@@

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|}
		\hline
		Platform name & Storage capacitor size \\
		\hline \hline
		Moo~\cite{moo} & 0.1\,F \\
		WISPCam~\cite{naderiparizi_rfid_2015} & 6.08\,mF \\ %tested [11.24, 17.45, 21.98]\,mF
		NFC-WISP~\cite{zhao_rfid_2015} & 300\,$\mu$F \\
		NeuralWISP~\cite{holleman_biocas_2008} & 100\,$\mu$F \\
		WISP~\cite{wisp5} & 47\,$\mu$F \\
		{\em BioImpedance} sensor~\cite{rodriguez_tbcs_2015} & 20\,$\mu$F \\
		{\em Ingestible} sensor~\cite{nadeau_naturebio_2017} & 220\,$\mu$F\\
		\hline
	\end{tabular} 
	\caption{Comparison of {\em default} energy storage sizes for various battery-less platforms. Observe a huge variation in storage capacity. We note that values for other representative platforms~\cite{medusa_farsens_2017,talla_imwut_2017,liu_sigcomm_2013,parks_sigcomm_2014} were not reported in their respective papers.}
	\label{table:capacitor}
\end{table}

@@

@@

Intermittent execution models combatting WAR can be classified into: \emph{sequential execution model} (which historically formed a baseline for this stream of research) and \emph{modular execution model} (a potent alternative, on which we base the design decisions of \sys).
)
\subsubsection{Baseline---Sequential Execution Model}

Under this model a program for intermittently-powered device is seen as a one idempotent operational region with one common context. Generally, the progress of the program is saved and updated by means of checkpointing, refer again to Section~\ref{sec:background_consistency}. Normally, the sequential model relies on a hardware assistant to measure the voltage level in the energy reservoir to place a checkpoint~\cite{mementos,mottola2017harvos,hibernus}. The benefit of this model is that it does not require code modification by a programmer. However it suffers from significant overhead (refer to experiment result in~\cite[Fig. 3]{chain}, where it is concluded that each checkpoint causes overhead). \todo{I use the word `checkpoint' for normal checkpoint and task boundary interchangeably---check if it does not cause confusion}{Brandon}

@@

At the heart of this model is the concept of an \emph{idempotent task}. The idempotent task is a C language function that does not have arguments and does not return a value. This task uses a well defined interface to interact with the non-volatile memory. Therefore, it tolerates arbitrary number of power interrupts. This model, generally, produces less overhead~\cite{chain} and allows multiple applications to run on an intermittently-powered device by interleaving their tasks. However, it obviously requires code modification. For example, if an algorithm is written according to the continuous execution model it has to be split by a programmer into small tasks to run under the modular intermittent execution model.

Despite the benefit of the modular model certain limitations are still present that must be overcome and which inspired us to design \sys. Let us look at these problems in detail.

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|c|c|}
		\hline
		{~} & CP & SZ & NV \\
		\hline\hline
		\!\!Mementos~\cite{mementos}\!\! & \!\!$N_2\leq N_1$\!\! & \!\!R + S\!\! & \!\!R + S\!\! \\
		\!\!DINO~\cite{dino}\!\! & $N_2$\!\! & \!\!R + W\!\! & \!\!R + W\!\! \\
		\!\!Chain~\cite{chain}, Alpaca~\cite{alpaca}\!\! & \!\!$N_2$\!\! & P\!\! & \!\!2W + P\!\!\\
		\!\!Ratchet~\cite{ratchet}, Clank~\cite{hicks_isca_2017}\!\! & $N_3\geq N_2\geq N_1$\!\! & \!\!R\!\! & R\!\!\\
		\hline 
	\end{tabular}
	\caption{Cost of various checkpointing methods; Symbols---R: register, S: stack, W: WAR-dependent variables, P: program counter, $N_1, N_2, N_3$: relative number of checkpoints for each method, CP: no. checkpoints, SZ: checkpoint size, NV: copy to non-volatile memory size.}
	\label{table:chechpoint_comparison}
\end{table}

@@

We have executed an experiment to measure the cost of read/write operations of $w$ to volatile and non-volatile memory. The result is given in Figure~\ref{fig:framEnergy}. Four applications were developed that \emph{access} the FRAM and SRAM of the MSP430FR5969 microcontroller~\cite{msp430datasheet}. The interference free debugger (EDB)~\cite{edb} is used as the energy measurement tool. The EDB probes the energy buffer before and after accessing the FRAM/SRAM 1600 times. This large number of FRAM access is used to increase the reliability of the results generated by the used measurement tool (e.g. reducing the effect of the quantization error). The energy buffer is charged, using the EDB, to $\approx$ 2.45\,V \textbf{only} at the beginning of the execution and of the programs. The write operations were performed by writing literal values, random numbers, to the memory. However, a read operation must be followed by a write operation. Therefore, we chose to write a read value to SRAM. Based on our measurements we conclude that a FRAM access equals $\frac{3}{2}$ SRAM access~\footnote{We remark that TI already mentions~\cite{ti_fram_faq} that FRAM write access is more energy expensive than SRAM write access. However, TI does not quantify this energy overhead.}. Therefore, reducing FRAM access is desirable when developing energy limited software solution---\emph{which is in sharp contrast to the main goal of all checkpoining solutions}.

The above result reassures the intuitive observation that a decrease of CP (Table~\ref{table:chechpoint_comparison}) decreases the energy cost associated with each state preservation method and reduces operation cost which in most simplistic terms is defined as $\mathcal{O}(\text{CP})+\mathcal{O}(\text{PL})$, where PL is the cost of program execution without checkpointing. On the other hand, this increases task/checkpoint re-execution time. The ideal solution is a runtime that \emph{adaptively} changes task size during runtime, which however exposes a second problem.

@@

A modular intermittent execution model completely depends on a programmer's estimation which mostly results in a sub-optimal code division. Moreover, this model can only consider a single hardware configuration and it does not take environment variability into considerations. Each battery-less platform has different storage sizes, refer to Table~\ref{table:capacitor}, which causes different charge/discharge periods. This result in different execution times. In the worst case the size of the capacitor might be too small to execute one task written by the programmer. This calls for the ability of the modular execution runtimes to \textbf{merge} multiple small tasks into a larger one or \textbf{split} a large task into a set of small ones on the fly (post-compilation time). Unfortunately the following features prevent us to perform this task using existing runtimes~\cite{chain,alpaca}.

@@

\begin{itemize}
	\item \textbf{Irreversible tasks transitions} According to both Chain~\cite{chain} and Alpaca~\cite{alpaca} programming models, each task calls the next one. The transition is done through functions calls and manually clearing the stack to prevent the stack overflow problem. This approach makes all the task transitions firm, preventing task merging.

	\item \textbf{Dependency on privatization} Alpaca~\cite{alpaca} extended Chain~\cite{chain} with a concept of privatization, i.e. it protects task-shared variables from WAR dependency problem \emph{within} the scope of a task. Therefore, if tasks are on-demand merged, the scope of these dependencies are changed and the static analysis to privatize these variables is not valid any more and the data consistency can not be preserved.

	\item \textbf{`All channels' requirement} Fig.~\ref{fig:DynamicChainSeq} shows that if the task control flow is a circular with no branches then task merging is possible. Moreover, all the intra-merged-tasks channel can be committed to SRAM instead of FRAM to save energy. However, this execution path is only a special one. If the general case (see again Fig.~\ref{fig:DynamicChainSeq}) is considered then leaving out the intra-merged-task channels might result in data inconsistency. Than is, if an application tries to access a merged task from the middle after a power interrupt then the input for that task is not ready, which result in an incorrect execution.

	\item \textbf{Lack of multi-versioning system} All known modular execution models define its own input for each task that is not shared with other tasks. This design decision makes the benefit of merging tasks of less importance since each task is interacting with its own variables. \todo{Rewrite this paragraph: still unclear}{Przemek}
\end{itemize}

\subsection{Main Idea: Tasks Coalescing}

To combat the above limitations we propose a new concept of \textbf{virtualization}: a modular execution model with softening a number of transitions, by keeping the state of the execution progress in the volatile memory, to construct a bigger \emph{virtual task} to reduce energy consumption. Ideally, the size of the virtual task should match the length of a continuous interval of the intermittent power supply for the best performance. Another form of virtualization can be achieved by reducing the number of non-volatile memory accesses. This can be done by having a temporary copy of a global variable in the volatile memory (similar to the caching principle) and a task interacts with it and updates it before writing back the most recent value to the non-volatile memory during the commit process.

Before proceeding further with the background we need to make certain assumptions about the underlying hardware. We assume that we operate on the microcontroller (more generally, computing hardware) equipped with a mix of volatile and non-volatile memory, such as FLASH, Ferroelectric RAM or Magnetoresistive RAM. We do not assume to rely on specific memory type (either volatile or non-volatile), in contrary to earlier works targeting non-volatile processors~\cite{su_date_2017,ratchet,quickrecall,nvp}. Actually, we will exploit the properties of both memory types to maximize performance metrics of the program being executed. We enable to I/O operations, however we do not enable the scheduling of operations (which \emph{nota bene} forms a foundation for a simple runtime). All in all, our hardware model follows in assumptions the two most relevant intermittently-powered execution environments~\cite{alpaca,chain}. \todo{Verify if this section is not out of place}{Brandon}

@@

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|c|c|}
		\hline
		{~} & Alpaca~\cite{alpaca} & Chain~\cite{chain} & This work \\
		\hline\hline
		Task connections & in-task & in-task & outside \\
		Task re-use & N & N & Y\\
		Memory model & globals & globals & globals\\
		Concurrency & N & N & N \\
		Task blocking & N & N & Y \\
		Energy-adaptive & N & N & Y \\
		\hline
	\end{tabular}
	\caption{Comparison of existing task-based execution models against \sys.}
	\label{table:feature_comparison}
\end{table}

@@
